### 2025-08-29 sections/5-discussion.tex
acknowleged unbalanced vs balanced section in discussion
back down to 10 pages with some extra material

We have presented CAKES, a suite of three algorithms for fast $k$-NN search across diverse distance functions.
When the distance is a metric (Sec.~\ref{sec:methods}), the algorithms are exact.
Under Cosine distance, a non-metric, they achieve nearly perfect recall.
CAKES is most effective when data satisfy the manifold hypothesis, i.e., they occupy a low-dimensional manifold within a high-dimensional space. 
It builds an unbalanced cluster tree that captures manifold structure and helps accelerate search.
Conversely, performance degrades on randomly distributed data, where such structure is absent.
In Figure~\ref{fig:results:lfd-plots}, we show the extent to which each dataset we tested exhibits a manifold structure, as quantified by the percentiles of LFD.

@@

We have noted how the fastest algorithm in CAKES varies depending on the dataset and distance function being used, as well as the cardinality of the dataset.
This variability motivated our auto-tuning process (Sec.~\ref{sec:methods:auto-tuning}).
Future work should examine why and when each algorithm excels and learn to use the geometric and topological properties of datasets to choose the ideal algorithm to use.

@@

Despite CAKES' generality, effective deployment on real data requires domain knowledge.
In protein sequence search, for example, sequences vary widely in length (typically 30-1000 amino acids, with outliers) and have a 20-letter alphabet.
Numerous algorithms exist~\cite{kim2021entrance, daniels2013compressive, yu2015entropy, steinegger2018clustering}, and their development relied on deep domain expertise.
Simple edit distance can be poor when lengths vary substantially.
CAKES can get around that issue by instead using alternative metrics or preprocessing—for example, represent each sequence as a set of $k$-mers and use Jaccard distance, as in~\cite{kim2021entrance}.
If the distance relies on global alignment (e.g., Needleman-Wunsch~\cite{needleman1970general}), the dataset can be partitioned into mutually exclusive length bins with a separate tree per bin;
at query time, we simply route the sequence to its bin by length, alogn with a small number of adjacent bins.

@@

CLAM and CAKES are implemented in Rust and available under the MIT license at https://github.com/URI-ABD/clam.

>>>>

We have presented CAKES, a suite of three algorithms for fast $k$-NN search across diverse distance functions.
When the distance is a metric (Sec.~\ref{sec:methods}), the algorithms are exact.
Under Cosine distance, a non-metric, they achieve nearly perfect recall.
CAKES is most effective when data satisfy the manifold hypothesis, i.e., they occupy a low-dimensional manifold within a high-dimensional space. 
It builds an unbalanced cluster tree that captures manifold structure and helps accelerate search. The results in Section~\ref{sec:results:clustering-strategies-and-number-of-distance-computations} shows that this unbalanced clustering approach is superior to balanced clustering in both absolute and asymptotic performance. 
Conversely, performance degrades on randomly distributed data, where such structure is absent.
In Figure~\ref{fig:results:lfd-plots}, we show the extent to which each dataset we tested exhibits a manifold structure, as quantified by the percentiles of LFD.

@@

We have noted how the fastest algorithm in CAKES changes depending on the dataset and distance function being used, as well as the cardinality of the dataset.
This variability motivated our auto-tuning process (Sec.~\ref{sec:methods:auto-tuning}).
Future work should examine why and when each algorithm excels and learn to use the geometric and topological properties of datasets to choose the ideal algorithm to use.

@@

Despite CAKES' generality, effective deployment on real data requires domain knowledge.
In protein sequence search, for example, sequences vary widely in length (typically 30-1000 amino acids, with outliers) and have a 20-letter alphabet.
Numerous algorithms exist~\cite{kim2021entrance, daniels2013compressive, yu2015entropy, steinegger2018clustering}, and their development relied on deep domain expertise.
Simple edit distance can be poor when lengths vary substantially.
CAKES can get around that issue by instead using alternative metrics or preprocessing—for example, the dataset can be partitioned into mutually exclusive length bins with a separate tree per bin;
at query time, we simply route the sequence to its bin by length, alogn with a small number of adjacent bins.

@@

CAKES is implemented in Rust and available under the MIT license at https://github.com/URI-ABD/clam.
